A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-20 tex/sec6-formalisation.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

\section{Formalisation} \label{sec:formalisation}
As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda.
While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could pose some difficulty.
If the reader is comfortable with the informal presentation and assured by the existence of its formalisation, they may feel free to skip this section.

@@

\paragraph{Category-theoretic analysis of well-typed terms}
The difference between intrinsic typing and extrinsic typing could be partly manifested by \citet{Fiore1999}{'s} theory of abstract syntax and variable binding which forms the foundation of \citet{Fiore2022}'s framework and inspires other referenced frameworks.
Let us take a closer look of their idea.
The set of (untyped) abstract syntax trees for a language can be understood as
\begin{enumerate*}
  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs. 
  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
\end{enumerate*}
Substitution for $\Term$ is modelled as a monoid in the category of presheaves.
To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.

@@

As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}

@@

The functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}} is defined for an operation $\biop$:
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[7]%
\>[7]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;\Conid{X})\;{}\<[E]%
\\
\>[7]{}\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}\;\mathrel{=}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{d}\;\Varid{≡}\;\Varid{d′}\;\Varid{×}\;\Conid{Σ[}\;\Varid{p}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{A₀}\;\Varid{⟨}\;\Varid{p}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
A set within the family is only inhabited if the indices correspond with the mode and the type as specified by the operation.
Therefore, in defining the set \ensuremath{\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}}, it takes identity proofs for \ensuremath{\Varid{d}\;\Varid{≡}\;\Varid{d′}} and \ensuremath{\Conid{A₀}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}}, with \ensuremath{\Conid{A₀}} instantiated by a substitution \ensuremath{\Varid{σ}} from $\Xi$ to $\emptyset$.
This ensures that the set is inhabited only if the indices align with what the arity (\ensuremath{\Conid{D}\;\Varid{.ar}\;\Varid{o}}) specifies.

@@

At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}}, similarly to how we have defined simple types and raw terms.
This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{56}{@{}>{\hspre}l<{\hspost}@{}}%
\column{62}{@{}>{\hspre}l<{\hspost}@{}}%
\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[32]%
\>[32]{}\hspace{10em}{}\<[56]%
\>[56]{}\Varid{\char95 ↑\char95 }\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
\>[56]{}\Varid{op}\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}

@@

@@

  \begin{minipage}[b]{.44\textwidth}

@@

@@

\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;{}\<[26]%
\>[26]{}\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[26]%
\>[26]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;{}\<[26]%
\>[26]{}\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Varid{op}\;\Varid{ts})\;{}\<[26]%
\>[26]{}\mathrel{=}\;{}\<[E]%

@@

\>[5]{}\mathbin{:}\;(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%

>>>>

\section{Formalisation} \label{sec:formalisation}
As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda.
While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could be difficult.
If the reader is comfortable with the informal presentation and assured by the existence of its formalisation, they may feel free to skip this section.

@@

\paragraph{Category-theoretic analysis of well-typed terms}
The difference between intrinsic typing and extrinsic typing could be partly manifested by \citet{Fiore1999}{'s} theory of abstract syntax and variable binding which forms the foundation of \citet{Fiore2022}'s framework and inspires other referenced frameworks.
Let us take a closer look of their idea.
The set of (untyped) abstract syntax trees for a language can be understood as
\begin{enumerate*}
  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs. 
  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
\end{enumerate*}
Substitution for $\Term$ is modelled as a monoid in the category of presheaves.
To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.

@@

As shown in \Cref{fig:formal-bibisig}, we define \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}

@@

The functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}} is defined for an operation $\biop$:
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[7]%
\>[7]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}\;\mathrel{=}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{d}\;\Varid{≡}\;\Varid{d′}\;\Varid{×}\;\Conid{Σ[}\;\Varid{p}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{A₀}\;\Varid{⟨}\;\Varid{p}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
A set within the family is only inhabited if the indices correspond with the mode and the type as specified by the operation.
Therefore, in defining the set \ensuremath{\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}}, it takes identity proofs for \ensuremath{\Varid{d}\;\Varid{≡}\;\Varid{d′}} and \ensuremath{\Conid{A₀}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}}, with \ensuremath{\Conid{A₀}} instantiated by a substitution \ensuremath{\Varid{σ}} from $\Xi$ to $\emptyset$.
This ensures that the set is inhabited only if the indices align with what the arity (\ensuremath{\Conid{D}\;\Varid{.ar}\;\Varid{o}}) specifies.

@@

At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}}, similarly to how we have defined simple types and raw terms.
This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{56}{@{}>{\hspre}l<{\hspost}@{}}%
\column{62}{@{}>{\hspre}l<{\hspost}@{}}%
\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
\column{85}{@{}>{\hspre}l<{\hspost}@{}}%
\column{89}{@{}>{\hspre}l<{\hspost}@{}}%
\column{108}{@{}>{\hspre}l<{\hspost}@{}}%
\column{113}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[32]%
\>[32]{}\hspace{10em}{}\<[56]%
\>[56]{}\Varid{\char95 ↑\char95 }\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇐}\;\Conid{A}\;{}\<[E]%
\\[\blanklineskip]%
\>[9]{}\Varid{\char95 ∋\char95 }\;{}\<[15]%
\>[15]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
\>[56]{}\Varid{op}\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;{}\<[85]%
\>[85]{}\Varid{rs}\;{}\<[89]%
\>[89]{}\phantom{[}{}\<[108]%
\>[108]{}\Varid{d}\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;{}\<[89]%
\>[89]{}[\mskip1.5mu \;{}\<[108]%
\>[108]{}\Varid{d}\;\mskip1.5mu]\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}

@@

As an illustration, consider the formal proof of soundness below, i.e.\ the `if' part of \cref{lem:soundness-completeness}.
Its formal proof is defined in a module which has signatures as its parameters:
\>[B]{}\Keyword{module}\;\Conid{Theory.Soundness}\;\{\mskip1.5mu \Conid{SD}\;\mathbin{:}\;\Conid{S.SigD}\mskip1.5mu\}\;(\Conid{BD}\;\mathbin{:}\;\Conid{B.SigD}\;\Conid{SD})\;\Keyword{where}{}\<[E]%
Then, the induction proof is nothing more than a function defined by pattern matching for each case: each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
(The definition \ensuremath{\Varid{soundnessᵃ}} might be further removed if \Agda's termination checker 
The only (slightly) interesting case $\ChkRule{Sub}$ is proved by disregarding the identity proof \ensuremath{\Varid{refl}} and use the induction hypothesis (\ensuremath{\Varid{soundness}\;\Varid{t}}) directly:
\begin{figure}[H]

@@

  \begin{minipage}[t]{.55\textwidth}

@@

\column{28}{@{}>{\hspre}l<{\hspost}@{}}%

@@

\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[24]%
\>[24]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%

@@

\>[3]{}\Varid{soundness}\;(\Varid{op}\;\{\mskip1.5mu \Varid{rs}\;\mathrel{=}\;\Varid{i}\;\Varid{,}\;\anonymous \mskip1.5mu\}\;(\anonymous \;\Varid{,}\;\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{ts}))\;\mathrel{=}\;{}\<[E]%

@@

\>[5]{}\mathbin{:}\;{}\<[8]%
\>[8]{}(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%
